Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/study.tex
various edits from sections 2-6

\begin{center}
\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
 \fortran{b(i) = a(i)  + a(i)} \\ \hline
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1,j) + a(i-1,j)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular}
\end{center}
%
%To assess the hypothesis that
%arrays are mostly read in a pattern that includes the immediate neighbours
%to the ``origin'', we categorised the position of subscript patterns for each
%right-hand side, in each dimension: \\
%
%\noindent
%
Finally, we categorise the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:
%
\begin{center}
\begin{tabular}{l|l|l}
  Shorthand & Classification & Example \\ \hline
  (\textsf{const}) & Consistent & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  (\textsf{perm}) & Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  (\textsf{LHSsub}) & LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  (\textsf{LHSsup}) & LHS superset & \fortran{a(i, j) = b(i)} \\
  (\textsf{inconst}) & Inconsistent & \fortran{a(i) = b(j)}
\end{tabular}
\end{center}
%
If an assignment is not inconsistent, but one of the other four
categories, we refer to it as *consistent.

@@

\noindent
We revisit each hypothesis and show the related data and conclusions.
%
\begin{enumerate}
\item Computations involving loops over arrays are common, where an
  arrays are read from with a static pattern based on constant offsets
  from (base or dervied) induction variables;
\end{enumerate}
%
We identified $108,773$ instances of assignments within loops in
which an array subscript flows to the right-hand side. We refer
to each one of these as an \emph{array computation}.
We performed the classification described above
and grouped the data by the right-hand side classification only:
%%
\begin{center}
\begin{tabular}{l|S[table-format=4]@{\extracolsep{5pt}}S[table-format=2.2]c}
\textbf{RHS classification} & \textbf{Number} & \% &
\textit{grouping 1} \\ \hline
Affine                          & 34        & 0.03\%
& \multirow{3}{*}{81.43\%} \\ \cline{1-3}
Neighbourhood                   & 65334     & 60.06 \%  \\ \cline{1-3}
Neighbourhood + absolute       & 23209     & 21.34\%  \\ \hline
Other                           & 20196     & 18.57 \%  \\ \hline \hline
\textbf{Total}                  & 108773    & 100\% \\
\end{tabular}
\end{center}
%
\noindent
Since, affine and neighbour patterns are static, we group them
(\textit{grouping 1}) to see that $81.43\%$ of the array computations
(88,577) have a static pattern that comprises constant offsets of
base/derived induction variables (potential with some absolute
indices), confirming \emph{hypothesis 1}. Note that affine subscripts
are rare, perhaps due to programmers relying on compilers for
optimisation rather than hand optimising (which tends to lead to
affine indices).  Overwhelmingly, the main category is neighbourhood
offsets (possibly mixed with constants in some dimensions). The
``affine + absolute'' class is not represented at all. \marginpar{\mnote{(2)
  Mention what sort of things are included in other.}}

\begin{enumerate}[resume]
\item Most loop-array computations of the previous form read
from arrays with a \emph{contiguous} pattern.
%
\item Most loop-array computations of the previous form read
from arrays with a pattern that includes the immediate
neighbours (offsets of 1 from the induction variables);
\end{enumerate}
%
We grouped the data based on subclassifications of \emph{grouping 1}
in terms of contiguity and positioning.
%Focussing on grouping 1.
\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]}
\textbf{RHS classification} & \textbf{Number} & \% {(of whole)}
&\textbf{Position} & \textbf{Number} & \% {(of whole)} \\ \hline
Affine, contiguous
& 31 & 0.03\% \\ \hline
%\multirow{3}{*}{Affine, contiguous}
% & %\multirow{3}{*}{31}
% & %\multirow{3}{*}{0.03\%}
% & \textsf{away} & 4 & 0.00\% \\
%& 31 & 0.03\% & \textsf{origin} & 15 & 0.01\% \\
%&&& \textsf{straddle} & 12 & 0.01\% \\ \hline
Affine, non-contiguous
 & 3
 & 0.00\% & \\ \hline
\multirow{3}{*}{Neighbour, contiguous}
 & %\multirow{3}{*}{83689}
 & %\multirow{3}{*}{83.08\%}
 & \textsf{away} & 152 & 0.14\% \\
& 83689 & 76.94\% & \textsf{origin} & 81309 & 74.75\% \\
&&& \textsf{straddle} & 2228 & 2.05\% \\ \hline
Neighbour, non-contiguous
 & 4854
 & 4.46\% \\ \hline
Other & 20196 & 18.57\%  \\
 \hline \hline
\textbf{Total}  & 108773    & 100\% \\
\end{tabular}
\end{center}
%%
Thus, $76.94\%$ of array computations have only neighbour
indices (which may include some absolute indices) in a contiguous
pattern on the right hand side, with $76.80\%$ either including the origin
($74.75\%$) or offset by a distance of 1 from the origin
($2.05\%$). This confirms both hypotheses 2 and 3. Only 152 instances
of index patterns not immediately next to or over the origin were
found.  There is a non trivial amount of non-contiguous arrays access,
but it is a much smaller proportion of the whole.

@@

We subclassified the data based on the right-hand side
and left-hand side categories

@@

 & \multirow{7}{*}{66.01\%} \\\cline{1-6}

@@

LHS IVs & RHS neigh   &  4560      & 4.19\%  & 50861     & 46.76\% \\\cline{1-6}

@@

LHS Vars & RHS affines & 0 & 0.00\%  & 4         & 0.00\% \\\cline{1-6}

@@

other       &                      & 30019     & 27.60\%  & &  \\\hline%\hline

@@

affine or a constant translation. Thus, the data supports ours

@@

Most stencils are contiguous, but there is a non-trivial
amount of non-contiguity. Most stencils have matching induction
variables on the left and right-hand sides but again there
are non-trivial amounts of the other kinds of relationship.
\item Many array computations read from each particular index just
  once.
We classified right-hand sides which are neighbourhood offsets (or
affine) and subclassified this by whether array subscripts were
unique or repeated:

\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]}
\textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
RHS neighbour/affine unique subscripts      & 78349     & 72.03\%  \\\hline
RHS neighbour/affine repeated subscripts   & 10228     & 9.40\%  \\\hline
other                             & 20196     & 18.57\%  \\\hline\hline
Total                             & 108773    &
\end{tabular}
\end{center}
%
Thus, hypothesis 5 is confirmed, the majority of array computations
have access patterns with linear usage of array subscripts.

>>>>

\begin{center}
\begin{tabular}{l||l|p{0.28\linewidth}|l}
  Property   & Shorthand & Classifications (of RHS pattern) & Example \\ \hline
%%%%%%
\multirow{2}{*}{Contiguity} & (\textsf{contig}) & Contiguous  &
                                                                \fortran{a(i) + a(i+1) + a(i+2)} \\
  & (\textsf{disjoint}) & Non-contiguous & \fortran{a(i) + a(i+2)} \\ \hline
%%%%%%
%\multirow{2}{*}{Shape} & (\textsf{rect}) & (Hyper)rectangle &
% \fortran{a(i,j) + a(i+1,j) + a(i,j+1) + a(i+1,j+1)} \\
%           & (\textsf{sumRect}) & Composed (hyper)rectangles &
%  \fortran{a(i,j) + a(i,j-1) + a(i-1,j) + a(i+1,j) + a(i,j+1)} \\ \hline
%%%%%%
  \multirow{2}{*}{Reuse} & (\textsf{readonce}) & Unique subscripts
                                         & \fortran{b(i) = a(i) + a(i+1)} \\
  & (\textsf{mult}) & Repeated subscripts &
 \fortran{b(i) = a(i)  + a(i)} \\ \hline
  \multirow{2}{*}{Positioning} & (\textsf{origin}) & Includes origin
& \fortran{a(i)} or \fortran{a(i+1, j)} (in $2^{\textit{nd}}$
  dimension) \\
             & (\textsf{straddle}) & Within distance 1 of origin &
\fortran{a(i+1,j) + a(i-1,j)} \\
             & (\textsf{away}) & Away from the origin
                                          & \fortran{a(i+2), a(i+3)}
\end{tabular}
\end{center}
%
%To assess the hypothesis that
%arrays are mostly read in a pattern that includes the immediate neighbours
%to the ``origin'', we categorised the position of subscript patterns for each
%right-hand side, in each dimension: \\
%
%\noindent
%
Finally, we categorise the relationship between the left-hand side
and right-hand side, in terms of the use of induction variables.
The two sides are \emph{consistent} if the same induction variables
appear in each side, used for the same dimension. This is
weakened to a \emph{permutation} if the roles of the induction
variables changes. This is weakened further if the left-hand side
induction variables are either
a subset or superset of the induction variables on the right-hand
side. Otherwise, the two sides are seen as inconsistent:
%
\begin{center}
\begin{tabular}{l|l}
  %Shorthand &
  Classification & Example \\ \hline
  %(\textsf{const}) &
  Consistent & \fortran{a(i, j) = b(i, j) +
                                  b(i+1,j+1)} \\
  %(\textsf{perm})
  Permutation & \fortran{a(i, j) = c(j, i)} or \fortran{a(i,
                                                               0) =
                                 b(0, i)} \\
  %(\textsf{LHSsub}) &
  LHS subset & \fortran{a(i) = b(i, j) + b(i, j-1)}
  \\
  %(\textsf{LHSsup}) &
  LHS superset & \fortran{a(i, j) = b(i)} \\
  %(\textsf{inconst}) &
  Inconsistent & \fortran{a(i) = b(j)}
\end{tabular}
\end{center}
%
If an assignment is not inconsistent, but is one of the other four
categories, we refer to it as *consistent.

@@

\noindent
We revisit each hypothesis and show the related data and conclusions.
%
\begin{enumerate}
\item Computations involving loops over arrays are common, where an
  arrays are read from with a static pattern based on constant offsets
  from (base or dervied) induction variables;
\end{enumerate}
%
We identified $108,773$ instances of assignments within loops in
which an array subscript flows to the right-hand side. We refer
to each one of these as an \emph{array computation}.
We performed the classification described above
and grouped the data by the right-hand side classification only:
%%
\begin{center}
\begin{tabular}{l|S[table-format=4]@{\extracolsep{5pt}}S[table-format=2.2]c}
\textbf{RHS classification} & \textbf{Number} & \% &
\textit{grouping 1} \\ \hline
Affine                          & 34        & 0.03\%
& \multirow{3}{*}{81.43\%} \\ \cline{1-3}
Neighbourhood                   & 65334     & 60.06 \%  \\ \cline{1-3}
Neighbourhood + absolute       & 23209     & 21.34\%  \\ \hline
Other                           & 20196     & 18.57 \%  \\ \hline \hline
\textbf{Total}                  & 108773    & 100.00\% \\
\end{tabular}
\end{center}
%
\noindent
Since, affine and neighbour patterns are static, we group them
(\textit{grouping 1}) to see that $81.43\%$ of the array computations
(88,577) have a static pattern that comprises constant offsets of
base/derived induction variables (potential with some absolute
indices), confirming \emph{hypothesis 1}. Note that affine subscripts
are rare, perhaps due to programmers relying on compilers for
optimisation rather than hand optimising (which tends to lead to
affine indices).  Overwhelmingly, the main category is neighbourhood
offsets (possibly mixed with constants in some dimensions). The
``affine + absolute'' class is not represented at all. %\marginpar{\mnote{(2)
%  Mention what sort of things are included in other.}}

\begin{enumerate}[resume]
\item Most loop-array computations of the previous form read
from arrays with a \emph{contiguous} pattern.
%
\item Most loop-array computations of the previous form read
from arrays with a pattern that includes the immediate
neighbours (offsets of 1 from the induction variables);
\end{enumerate}
%
We grouped the data based on subclassifications of \emph{grouping 1}
in terms of contiguity and positioning.
%Focussing on grouping 1.
\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]lS[table-format=4]S[table-format=2.2]}
\textbf{RHS classification} & \textbf{Number} & \% {(of whole)}
&\textbf{Position} & \textbf{Number} & \% {(of whole)} \\ \hline
Affine, contiguous
& 31 & 0.03\% \\ \hline
%\multirow{3}{*}{Affine, contiguous}
% & %\multirow{3}{*}{31}
% & %\multirow{3}{*}{0.03\%}
% & \textsf{away} & 4 & 0.00\% \\
%& 31 & 0.03\% & \textsf{origin} & 15 & 0.01\% \\
%&&& \textsf{straddle} & 12 & 0.01\% \\ \hline
Affine, non-contiguous
 & 3
 & 0.00\% & \\ \hline
\multirow{3}{*}{Neighbour, contiguous}
 & %\multirow{3}{*}{83689}
 & %\multirow{3}{*}{83.08\%}
 & \textsf{away} & 152 & 0.14\% \\
& 83689 & 76.94\% & \textsf{origin} & 81309 & 74.75\% \\
&&& \textsf{straddle} & 2228 & 2.05\% \\ \hline
Neighbour, non-contiguous
 & 4854
 & 4.46\% \\ \hline
Other & 20196 & 18.57\%  \\
 \hline \hline
\textbf{Total}  & 108773    & 100.00\% \\
\end{tabular}
\end{center}
%%
Thus, $76.94\%$ of array computations have only neighbour
indices (which may include some absolute indices) in a contiguous
pattern on the right hand side, with $76.80\%$ either including the origin
($74.75\%$) or offset by a distance of 1 from the origin
($2.05\%$). This confirms both hypotheses 2 and 3. Only 152 instances
of index patterns not immediately next to or over the origin were
found.  There is a non trivial amount of non-contiguous arrays access,
but it is a much smaller proportion of the whole.

@@

We subclassified our data by r.h.s
and l.h.s categories

@@

         & \\\hline
  LHS Vars & RHS affines & 0 & 0.00\%  & 4         & 0.00\% & 
  \\\hline

@@

LHS IVs & RHS neigh   &  4560      & 4.19\%  & 50861     & 46.76\% &
                                                                     \multirow{4}{*}{61.79\%}
  \\\cline{1-6}

@@

@@

other       &                      & 30019     & 27.60\%  & &
  \\\hline\hline
 Total & & 36975 & 34.00\% & 71798 & 66.01\%

@@

a neighbour or affine, and $61.79\%$ we classify as \emph{stencils},
writing to an array subscript on the left. Thus, the data supports ours

@@

The additional column in the right table (relativised) counts the
number of specifications where the left-hand side has neighbourhood
offsets, \eg{}, \fortran{a(i+1) = b(i)}.

Most stencils are contiguous, but there is some
non-contiguity. Most stencils have matching induction
variables on the left and right-hand sides but again there
are non-trivial amounts of the other kinds of relationship.
%%
\begin{enumerate}[resume]
\item Many array computations read from each particular index just
  once.
\end{enumerate}
%
We classified right-hand sides comprising neighbourhood (or
affine) subscripts and subclassified this by whether array subscript terms were
all unique or whether there were some repeats.
%
\begin{center}
\begin{tabular}{l|S[table-format=4]S[table-format=2.2]S[table-format=4]}
\textbf{RHS form and reuse} & \textbf{Number} & \% \\ \hline
RHS neighbour/affine unique subscripts      & 78349     & 72.03\%  \\\hline
RHS neighbour/affine repeated subscripts   & 10228     & 9.40\%  \\\hline
other                             & 20196     & 18.57\%  %\\\hline\hline
%Total                             & 108773    &
\end{tabular}
\end{center}
%
Thus most array computations have access patterns with
unique use of array subscripts, confirming hypothesis 5.

\subsection{Additional properties of array computations}
\label{subsec:additional-data}

\noindent
For each of the $108,773$ array computations, we recorded four
additional properties: the max dimensionality of array
subscripts involved, %the maximum depth of neighbourhood offsets,
the number of subscript terms involved in the computation, and the
length of the dataflow path (\ie{}, how many intermediate assignments
contribute to the array computation). The following gives the data
as histograms:

\setlength{\tabcolsep}{0.25em}
\begin{center}
\begin{tabular}{l|ccccccccccc}
  \textbf{Dimensionality} & 0 & 1     & 2     & 3     & 4    & 5   & 6 & 7 & 8 & 9 & 10 \\\hline
  Count                   & 0 & 43,227 & 32,830 & 24,621 & 7,608 & 467 & 13 & 5 & 0 & 0 & 3
  \end{tabular}
\end{center}

%\begin{center} \begin{tabular}{l|ccccccccccccccccccccccc}
%   \textbf{Max. offset depth}: & 0      & 1    & 2   & 3  & 4  & 5
%    & 6  & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & \ldots & 21 & 22 &
%    \ldots & 45 & \ldots & 100 \\\hline
%    Count & 207,512 & 9,530 & 327 & 57 & 55 & 7 & 11 & 0 & 0 & 1 & 11 &
%    1 & 8  & 4  & 0  & 10 & \ldots & 1  & 1  & \ldots & 8  & \ldots & 2
%   \end{tabular}
% \end{center}

   \begin{center}     \begin{tabular}{l|ccccccccccccccccccccc}
   \textbf{\# subscripts}: & 0 & 1     & 2     & 3    &
                                                                       4
     & 5   & 6   & 7   & 8   & 9   & 10  & 11 & 12 & 13 & 14 & 15 & 16
     & 17 & 18 & 19 & 20$\leq$ \\\hline
   Count & 0 & 84,690 & 14,377 & 4,234 & 2,367 & 881 & 607 & 192 & 457 & 284 & 207 & 45 & 72 & 33 & 16 & 41 & 40 & 27 & 15 & 4  & 184
   \end{tabular}
\end{center}

 \begin{center}
\begin{tabular}{l|ccccccccccccccccccccc}
  \textbf{Dataflow length}: & 0      & 1     & 2    & 3
  & 4    & 5 & 6    & 7   & 8   & 9   & 10  & 11  & 12  & 13  & 14
  & 15$\leq$ \\\hline
  Count & 106,712 & 13,323 & 7,091 & 3,898 & 3,031 & 1,695 & 1,333 & 965 & 792 & 704 & 482 & 430 & 339 & 310 & 270 & 2544
\end{tabular}
\end{center}
%
The histogram of ``number of subscripts'' shows that most
computations are relatively simple, with only 1-3 array subscripts.
However, some are complex, with 184 with over 20 indexing terms. We had one example with 96
contributing array subscripts. Thus, we see that there are many cases
in practice with complex patterns, that we aim to provide support for
with our specification language.

The rest of the data here is of use later when considering the
performance of our checking procedure (\Cref{subsec:checking}).

\subsection{From experiment to language design}

\noindent
From the data,
we make the following design decisions:
%
\begin{itemize}[leftmargin=1.5em]
\item The core support should be for neighbourhood offsets on
right-hand side of assignments in loops ($81.43\%$
of our array computations);

\item Access patterns are mostly contiguous ($76.94\%$) and cross or
straddle the origin ($75.75\%$), thus we can represent patterns by
\emph{finite intervals} that meet at, or are next to, the origin.
These intervals will be given declaratively and abstractly, avoiding
replicating indexing terms at the specification level.

\item We will support only consistent left and right-hand sides
(where there are common induction variables on each side) ($66.01\%$)

\item Stencils will be the primary focus $(61.79\%)$ but we will also
allow specifications on assignments where the l.h.s is a variable
(\eg{}, in a reduction) ($4.22\%$).

\item There is some non-contiguity ($4.46\%$) which we will specify
by finite intervals given as approximations that cover a non-contiguous pattern.

\item Affine indexing is very rare $<0.03\%$ so we do not consider it.
\end{itemize}
%
